Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-03-03 manuscript/climate_sensitivity_draft.Rmd
finished incorporating Bill's & Alan's comments (...wjm_AT)

We defined years with widespread growth reduction ("pointer years") as those where >25% of the cored trees experienced >30% reduction in basal area increment ($BAI$) relative to the previous 5 years, following the drought resistance ($Rt$) metric of [@lloret_components_2011]. $Rt$ values <1 indicate growth reductions, whereas values >1.0 indicate increased growth. Pointer years were identified using the pointRes package [@R-pointRes] in R. Four years met our criteria: 1966, 1977, 1991, and 1999. We excluded 1991 (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not identified as among the driest of the time period (Table S3). Rather, the severity of growth reduction may be explained in part by defoliation from gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which strongly impacted *Quercus* spp. [@twery_effects_1991].

Together, these criteria identified three drought years: 1966, 1977, and 1999 (Figs. 1, S2, Table S3). The droughts differed in intensity and prior onset (Fig. S2, Table S3). 
The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall and in August reached the minimum growing season $PDSI$ (-4.82) among of any of the three droughts. 
The 1977 drought was the least intense throughout the growing season, and was preceded by 2.5 years of near-normal conditions, making it the mildest of the three droughts. 
The 1999 drought was preceded by wetter than average conditions until the previous June, but reached the lowest PDSI during May-July.

@@

For each drought period, we calculated drought resistance ($Rt$) as the ratio of BAI during drought to the mean BAI over the five years preceding the drought [@lloret_components_2011]. Thus, $Rt$<1 indicates reduced growth under drought. Analyses focused on testing the predictions presented in Table 1, with $Rt$ as the response variable. The general statistical model for hypothesis testing was a mixed effects model (lme4 package from @R-lme4) with $Rt$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects. We used AICc (AICcmodavg package from @R-AICcmodavg) to assess model selection, and conditional/marginal R-squared to assess model fit.

Models were run for all drought years combined (with year as a fixed effect) and for each drought year independently. In order to determine the relative importance of the traits alone, we first tested the predictor variables independently against both height and $Rt$ given height's substantial influence. Variables were considered to have significant influence on $Rt$ when AICc was reduced by $\ge$ 2 relative to the corresponding null model lacking that variable (Table 4).

We then determined the best full models for predicting $Rt$ for each individual drought year and for all years combined. Candidate variables were selected, based on the single-variable tests, as those whose addition to a corresponding null model improved fit (at dAICc $\ge$ 1.0) in at least one drought (Table 4). We compared models with all possible combinations of candidate variables and identified the full set of models within dAICc=1 of the very top model (that with lowest AICc), henceforth referred to as "full models". When a variable appeared in all top models and the sign of the coefficient was consistent across models, this was counted as support for/ rejection of the associated prediction by the full models. If the variable appeared in only some of the models, we considered this partial support/rejection.

All analysis beyond basic data collection was performed using R version 3.5.3 [@R-base]. All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI- ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and [TBD], respectively).

@@

Community-level tree growth responses to all three droughts were modest, with mean $Rt$ values of 0.86, 0.84, and 0.86 for 1966, 1977, and 1999 droughts, respectively (Fig. 1b). In each drought, roughly 30% of the cored trees experienced $\ge$ 30% growth reductions ($Rt \le 0.7$): 29% in 1966, 32% in 1977, and 27% in 1999. However, some individuals exhibited increased growth: ($Rt > 1.0$): 26% in 1966, 22% in 1977, and 26% in 1999.

@@

Larger-diameter trees showed greater reductions in growth during drought, although there was no significant effect during 1977 or 1999 individually (Tables 1, 4). The same held true for $ln[H]$ in single-variable tests (Tables 1, 4). When combined with other predictor variables in the full models, $ln[H]$ appeared, with negative coefficient, in all full models for the three droughts combined, in the 1966 model, and in one of the two 1999 models (Tables 1, 5).

Crown position varied as expected with $H$ (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig. 2d). Crown position was a much poorer predictor of $Rt$ than was $H$ in the single-variable tests (Table 4), lending little overall support to the hypothesis that trees with more exposed crowns have lower $Rt$ (Table 1). When considered alone, $CP$ had a significant response only in the 1966 drought, during which trees with dominant $CP$ had the lowest $Rt$. When $H$ was included in the model, $CP$ was a significant predictor in the 1999 drought, with lowest $Rt$ for suppressed and then intermediate trees. Crown position was included in some of the full models (Table 5). In 1977, where $H$ was not included in the full model, dominant trees had the lowest $Rt$, and suppressed the highest. In contrast, in full models including both $H$ and $CP$ (all droughts and 1999), the lowest $Rt$ was in suppressed, followed by intermediate, trees.

In the non-drought years for which we have vertical profiles in climate data (2016-2018), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig. 2). Specifically, maximum daily wind speeds were significantly higher above the top of the canopy (40-50m) than within and below (10-30m) (Fig. 2a). Relative humidity was also somewhat lower during June-August, ranging from ~50-80% above the canopy and ~60-90% in the understory (Fig. 2b). Air temperature did not vary across the vertical profile (Fig. 2c).

@@

Hydraulic traits, including $XP$, $PLA_{dry}$, and $\pi_{tlp}$, were linked to drought responses (Tables 1,4,5). In the single-variable tests, $LMA$ and $WD$ never significantly associated with $Rt$ (Table 4) and were excluded from the full models. In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4). $PLA_{dry}$ was a strong predictor for the 1966 drought and all droughts combined, with consistently negative coefficients (Table 4). Similarly, $PLA_{dry}$ was consistently included, with negative coefficient, in full models for the three droughts combined and for the 1966 and 1977 droughts individually (Table 5). $\pi_{tlp}$ did not come out significant in any single-variable tests; however, coefficients were consistently negative (Table 4) and $\pi_{tlp}$ was included in the top full model for all droughts combined and for the 1977 and 1999 droughts individually (Table 5). Xylem porosity was not significant for all droughts combined and had contrasting effects in the individual droughts: whereas ring-porous species had higher $Rt$ than diffuse- and semi-ring- porous species in the 1966 and 1999 droughts, they had lower $Rt$ in 1977 (Tables 4,5).

@@

Tree size, microenvironment, and hydraulic traits shaped tree growth responses across three droughts (Table 1). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@liu_effect_1993; @stovall_tree_2019]. There was a marginal additional effect of crown exposure, with a tendency for lowest $Rt$ among the most exposed (dominant) and suppressed trees. There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ (consistent with @zuleta_drought-induced_2017; @stovall_tree_2019), and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. The negative effect of height on $Rt$ held when also accounting for species' traits.  Drought sensitivity was not consistently linked to species' $LMA$, $WD$, or xylem architecture, but was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019]. This is a novel finding in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of responses was mostly consistent across droughts, supporting the conclusion that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Tables 4-5), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings significantly advance our knowledge of the factors that confer vulnerability or resistance on trees during drought.

The droughts considered here were of a magnitude that has occurred with an average frequency of approximately one per 10-15 years (Fig. 1a, @helcoski_growing_2019) and had modest impacts on tree growth (Fig. 1b). These droughts were classified as severe (1977) or extreme (1966, 1999) according to the PDSI metric and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019]; however, extreme, multiannual droughts ("megadroughts") of the type that have triggered massive tree die-off in other regions (e.g., @allen_global_2010; @stovall_tree_2019) have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016]. Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig. S2), severely stressed a larger portion of trees (Fig. 1b). It may be notable that the tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this tendency increases with drought strength [@bennett_larger_2015; @stovall_tree_2019]. Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig. 1b), potentially due to decreased leaf area of competitors during the drought. It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5).

@@

Our analysis has the limitation that canopy positions were recorded in 2018, as opposed to the years of the droughts. However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing canopy positions would bias against the acceptance of our hypothesis. The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees [@suarez_factors_2004; @scharnweber_confessions_2019]. It is safe to assume that currently suppressed trees have always been in the understory, and their relatively low $Rt$ (after accounting for height effects) is real, perhaps as a result of competition (*[Sohn et al. 2016](https://www.sciencedirect.com/science/article/pii/S0378112716304029)*). The observed height-sensitivity of $Rt$, together with the lack of advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests. Additional research comparing drought responses of young and old forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.

The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] gave us the sample size to compare historical drought responses across species and associated traits at a single site [see also @elliott_forest_2015]. Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response. The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses of tree growth (Tables 1,4,5) is both novel and consistent with previous studies linking these traits to habitat and drought tolerance. Previous studies have demonstrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019; @simeone_coupled_2019; @marechaux_drought_2015] (*Rosas et al. 2019, DOI: 10.1111/nph.15684*, [Fletcher et al. 2018](https://bsapubs.onlinelibrary.wiley.com/doi/full/10.1002/ajb2.1164)), and our findings indicate that these traits also influence drought responses. Furthermore, the observed linkage of $\pi_{tlp}$ to $Rt$ in this forest aligns with observations in the Amazon that $\pi_{tlp}$ is higher in drought-intolerant that drought-tolerant plant functional types and adds support to the idea that this trait is useful for categorizing and representing species' drought responses in models (*Powell et al. 2016, DOI: 10.1111/gcb.13731*). Because both $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily [@bartlett_rapid_2012; @scoffoni_leaf_2014], they hold promise for predicting drought growth responses across species. The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that don't form annual rings, and for diverse tropical forests. Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (*Powell et al. 2016, DOI: 10.1111/gcb.13731*).

As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here. Large trees have been disproportionately impacted in forests around the world [@bennett_larger_2015; @stovall_tree_2019], and we show, at least at this site, that this is primarily driven by their height with some contributions from canopy position. The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas solitary trees or the dominant trees in young regrowth forests should be less vulnerable. This would suggest that, all else being equal, mature forests would be more vulnerable to drought than young forests with short trees; however, root water access may limit the young forests [@bretfeld_plant_2018], and species traits often shift as forests age. Early successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) display a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for addressing how drought tolerance changes as forests age [e.g. @rodriguez-caton_long-term_2015]. In the meantime, the results of this study advance our knowledge of the factors conferring drought vulnerability and resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.

### Acknowledgements 
We especially thank the numerous researchers who helped to collect the data used here, in particular Jennifer C. McGarvey, Jonathan R. Thomspon, and Victoria Meakem for original collection and processing of cores. Thanks also to Camila Medeiros for guidance on hydraulic and functional trait measurements, Edward Brzostek's lab for collaboration on leaf sampling, and Maya Prestipino for data collection. Funding for the establishment of the SCBI ForestGEO Large Forest Dynamics Plot was provided by the Smithsonian-led Forest Global Earth Observatory (ForestGEO), the Smithsonian Institution, and the HSBC Climate Partnership. This study was funded by ForestGEO, a Virginia Native Plant Society grant to KAT and AJT, and support from the Harvard Forest and National Science Foundation which supports the PalEON project (NSF EF-1241930) for NP.

### Author Contribution 
KAT, IM, and AT designed the research. Tree-ring chronologies were developed by RH under guidance of AT and NP. Trait data was collected by IM, JZ under guidance of NK and LS. Other plot data were collected by IM, AS, EGA, and NB under guidance of EGA and WM. Data analyses were performed by IM under guidance of KAT and VH. KAT and IM interpreted the results. IM and KAT wrote the first draft of manuscript, and all authors contributed to revisions.

>>>>

We defined years with widespread growth reduction ("pointer years") as those where >25% of the cored trees experienced >30% reduction in basal area increment ($BAI$) relative to the previous 5 years, following the drought resistance ($Rt$) metric of [@lloret_components_2011]. $Rt$ values <1 indicate growth reductions, whereas values >1.0 indicate above-average growth. Pointer years were identified using the pointRes package [@R-pointRes] in R. Four years met our criteria: 1966, 1977, 1991, and 1999. We excluded 1991 (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not identified as among the driest of the time period (Table S3). Rather, the severity of growth reduction could probably be explained in part by defoliation by the gypsy moth (*Lymantria dispar* L.), which was documented to have strongly impacted *Quercus* spp. in the area from approximately 1988 through 1995 [@twery_effects_1991].

Together, these criteria identified three drought years: 1966, 1977, and 1999 (Figs. 1, S2, Table S3). The droughts differed in intensity and antecedent moisture conditions (Fig. S2, Table S3). 
The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall and in August reached the lowest growing season $PDSI$ (-4.82) of the three droughts. 
The 1977 drought was the least intense throughout the growing season, and it was preceded by 2.5 years of near-normal conditions, making it the mildest of the three droughts. 
The 1999 drought was preceded by wetter than average conditions until the previous June, but reached the lowest PDSI during May-July (-4.53).

@@

For each drought year, we calculated drought resistance ($Rt$) as the ratio of BAI during drought to the mean BAI over the five years preceding the drought [@lloret_components_2011]. Thus, $Rt$<1 indicates reduced growth under drought. Analyses focused on testing the predictions presented in Table 1, with $Rt$ as the response variable. The general statistical model for hypothesis testing was a mixed effects model (lme4 package from @R-lme4) with $Rt$ as the response variable, tree nested within species as a random effect, and one or more independent variables as fixed effects. We used AICc (AICcmodavg package from @R-AICcmodavg) to assess model selection, and conditional/marginal R-squared to assess model fit.

Models were run for all drought years combined (with year as a fixed effect) and for each drought year individually. In order to determine the relative importance of the traits alone, we first tested the predictor variables independently against both height and $Rt$ given height's substantial influence. Variables were considered to have significant influence on $Rt$ when AICc was reduced by $\ge$ 2 units relative to the corresponding null model lacking that variable (Table 4).

We then determined the best full models for predicting $Rt$ for each individual drought year and for all years combined. Candidate variables were selected, based on the single-variable tests, as those whose addition to a corresponding null model improved fit (at dAICc $\ge$ 1.0) in at least one drought year (Table 4). We compared models with all possible combinations of candidate variables and identified the full set of models within dAICc=1 of the very top model (that with lowest AICc), henceforth referred to as "full models". When a variable appeared in all top models and the sign of the coefficient was consistent across models, we viewed this as support for the acceptance/rejection of the associated prediction by the full models. If the variable appeared in only some of the models, we considered this partial support/rejection.

All analysis beyond basic data collection was performed using R version 3.5.3 [@R-base]. All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and [TBD], respectively).

@@

Community-level tree growth responses to all three droughts were modest, with a mean $Rt$ of 0.86 in 1966 and 1999, and 0.84 in 1977  (Fig. 1b). In each drought, roughly 30% of the cored trees experienced $\ge$ 30% growth reductions ($Rt \le 0.7$): 29% in 1966, 32% in 1977, and 27% in 1999. However, some individuals exhibited increased growth: ($Rt > 1.0$): 26% in 1966, 22% in 1977, and 26% in 1999.

@@

Larger-diameter trees showed greater reductions in growth when evaluating the three drought years together and for 1966 individually, although DBH was not significant during 1977 or 1999 individually (Tables 1, 4). The same held true for $ln[H]$ in single-variable tests (Tables 1, 4). When combined with other predictor variables in the full models, $ln[H]$ appeared, with negative coefficient, in all full models for the three droughts combined, in the 1966 model, and in one of the two models for 1999 (Tables 1, 5).

Crown position varied as expected with $H$ (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig. 2d). Crown position was a much poorer predictor of $Rt$ than was $H$ in the single-variable tests (Table 4), lending little overall support to the hypothesis that crown exposure reduces $Rt$ (Table 1). When considered alone, $CP$ had a significant response only in the 1966 drought, during which trees with dominant $CP$ had the lowest $Rt$. When $H$ was included in the model, $CP$ was a significant predictor in the 1999 drought, with lowest $Rt$ for suppressed and then intermediate trees. Crown position was included in some of the full models (Table 5). In 1977, where $H$ was not included in the full model, dominant trees had the lowest $Rt$, and suppressed the highest. In contrast, in full models including both $H$ and $CP$ (all droughts and 1999), the lowest $Rt$ was in suppressed, followed by intermediate, trees.

In the years for which we have vertical profiles in climate data (2016-2018), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig. 2). Specifically, maximum daily wind speeds were significantly higher above the top of the canopy (40-50m) than within and below (10-30m) (Fig. 2a). Relative humidity was also somewhat lower during June-August, ranging from ~50-80% above the canopy and ~60-90% in the understory (Fig. 2b). Air temperature did not vary across the vertical profile (Fig. 2c).

@@

Hydraulic traits, including xylem porosity ($XP$), $PLA_{dry}$, and $\pi_{tlp}$, were linked to drought responses (Tables 1,4,5). In the single-variable tests, $LMA$ and $WD$ were never significantly associated with $Rt$ (Table 4) and were excluded from the full models. In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation (dAIC > 1.0) during at least one of the three droughts (Table 4). $PLA_{dry}$ was a strong predictor for 1966 and all droughts combined, with consistently negative coefficients (Table 4). Similarly, $PLA_{dry}$ was consistently included, with negative coefficient, in full models for the three droughts combined and for the 1966 and 1977 droughts individually (Table 5). $\pi_{tlp}$ was not significant in any single-variable tests; however, coefficients were consistently negative (Table 4) and $\pi_{tlp}$ was included in the top full model for all droughts combined and for the 1977 and 1999 droughts individually (Table 5). Xylem porosity was not significant for all droughts combined and had contrasting effects in the individual droughts: whereas ring-porous species had higher $Rt$ than diffuse- and semi-ring- porous species in the 1966 and 1999 droughts, they had lower $Rt$ in 1977 (Tables 4,5).

@@

Tree size, microenvironment, and hydraulic traits shaped tree growth responses across three droughts (Table 1). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height rather than crown exposure [@liu_effect_1993; @stovall_tree_2019]. We found only a marginal additional effect of crown exposure, with a tendency for lowest $Rt$ among the most exposed (dominant) and suppressed trees. There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ (consistent with @zuleta_drought-induced_2017; @stovall_tree_2019), and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. The negative effect of height on $Rt$ held after accounting for species' traits.  Drought sensitivity was not consistently linked to species' $LMA$, $WD$, or xylem type (ring- vs. diffuse porous), but was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts. This is a novel finding in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of responses was mostly consistent across droughts, supporting the conclusion that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Tables 4-5), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought--and, by extension, likely make them more vulnerable to mortality (**Sapes et al. 2019, doi: 10.1093/treephys/tpz062**).

The droughts considered here were of a magnitude that has occurred with an average frequency of approximately once every 10-15 years (Fig. 1a, @helcoski_growing_2019) and had modest impacts on tree growth (Fig. 1b). These droughts were classified as severe (1977) or extreme (1966, 1999) according to the PDSI metric and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019]; however, extreme, multiannual droughts ("megadroughts") of the type that have triggered massive tree die-off in other regions (e.g., @allen_global_2010; @stovall_tree_2019) have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016]. Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig. S2), severely stressed a larger portion of trees (Fig. 1b). It may be notable that the tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this tendency increases with drought severity [@bennett_larger_2015; @stovall_tree_2019]. Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig. 1b), potentially due to decreased leaf area of competitors during the drought. It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5).

@@

Our analysis has the limitation that canopy positions were recorded in 2018, as opposed to the years of the droughts. However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing canopy positions would bias against the acceptance of our hypothesis. The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing greater drought sensitivity in more exposed trees [@suarez_factors_2004; @scharnweber_confessions_2019]. It is safe to assume that currently suppressed trees been suppressed throughout our analysis period, and their relatively low $Rt$ (after accounting for height effects) is real, perhaps as a result of competition (*[Sohn et al. 2016](https://www.sciencedirect.com/science/article/pii/S0378112716304029)*). The observed height-sensitivity of $Rt$, together with the lack of advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests. Additional research comparing drought responses of young and old forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.

The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] gave us the sample size to compare historical drought responses across species and associated traits at a single site [see also @elliott_forest_2015]. Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response. The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses of tree growth (Tables 1,4,5) is both novel and consistent with previous studies linking these traits to habitat and drought tolerance. Previous studies have demonstrated that $\pi_{tlp}$ and $PLA_{dry}$ are physiologically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019; @simeone_coupled_2019; @marechaux_drought_2015] (*Rosas et al. 2019, DOI: 10.1111/nph.15684*, [Fletcher et al. 2018](https://bsapubs.onlinelibrary.wiley.com/doi/full/10.1002/ajb2.1164)), and our findings indicate that these traits also influence drought responses. Furthermore, the observed linkage of $\pi_{tlp}$ to $Rt$ in this forest aligns with observations in the Amazon that $\pi_{tlp}$ is higher in drought-intolerant than drought-tolerant plant functional types and adds support to the idea that this trait is useful for categorizing and representing species' drought responses in models (*Powell et al. 2016, DOI: 10.1111/gcb.13731*). Because both $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily [@bartlett_rapid_2012; @scoffoni_leaf_2014], they hold promise for predicting drought growth responses across species. The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that don't form annual rings, and for diverse tropical forests. Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage hydraulic of traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (*Powell et al. 2016, DOI: 10.1111/gcb.13731*).

As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here. Large trees have been disproportionately impacted in forests around the world [@bennett_larger_2015; @stovall_tree_2019], and we show, at least at this site, that this is primarily driven by their height with some contributions from canopy position. The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas shorter solitary trees or the dominant trees in young forests that recently established after logging or natural disturbances should be less vulnerable. This would suggest that, all else being equal, mature forests would be more vulnerable to drought than young forests with short trees; however, root water access may limit the young forests [@bretfeld_plant_2018], and species traits often shift as forests age. Early- to mid- successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) display a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for addressing how drought tolerance changes as forests age [e.g. @rodriguez-caton_long-term_2015]. In the meantime, the results of this study advance our knowledge of the factors conferring drought vulnerability and resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.

### Acknowledgements 
We especially thank the numerous researchers who helped to collect the data used here, in particular Jennifer C. McGarvey, Jonathan R. Thompson, and Victoria Meakem for original collection and processing of cores. Thanks also to Camila Medeiros for guidance on hydraulic and functional trait measurements, Edward Brzostek's lab for collaboration on leaf sampling, and Maya Prestipino for data collection. Funding for the establishment of the SCBI ForestGEO Large Forest Dynamics Plot was provided by the Smithsonian-led Forest Global Earth Observatory (ForestGEO), the Smithsonian Institution, and the HSBC Climate Partnership. This study was funded by ForestGEO, a Virginia Native Plant Society grant to KAT and AJT, and support from the Harvard Forest and National Science Foundation which supports the PalEON project (NSF EF-1241930) for NP.

### Author Contribution 
KAT, IM, and AJT designed the research. Tree-ring chronologies were developed by RH under guidance of AJT and NP. Trait data was collected by IM, JZ under guidance of NK and LS. Other plot data were collected by IM, AS, EGA, and NB under guidance of EGA and WM. Data analyses were performed by IM under guidance of KAT and VH. KAT and IM interpreted the results. IM and KAT wrote the first draft of manuscript, and all authors contributed to revisions.
